Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-03 doc/invt.tex
invt.tex: add asymptotic limits

In this study,
we shall give another proof of the inverse-time formula.
%
We then adapt the formula
to updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
We shall show that the optimal updating magnitude
depends nontrivially on the simulation length
and other parameters.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is best in asymptotic convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

%\subsubsection{\label{twolimits}
%Two limits of the optimal schedule, $\alpha(t)$}
%
%
%The optimal $\alpha(t)$ that minimizes the error
%Eq. \eqref{eq:error_split}
%is still difficult to find analytically.
%%
%However, we can study two limiting cases,
%for the asymptotic part, Eq. \eqref{eq:error_asym1}.
%
%For $\tau \ll t$,
%the dominant contribution comes from
%the term(s) with the smallest eigenvalue $\lambda_k = \lambda_n$,
%and
%$$
%\dot u_n(\tau) = c,
%\quad
%\mathrm{and}\;
%u_n(\tau) = c \, (\tau + t_0),
%$$
%which means
%$$
%\alpha(\tau) \approx \frac{1}{\lambda_n \, (\tau + t_0)}.
%$$
%
%
%For $\tau \approx t$,
%$u_k(q) \approx 1$,

>>>>

In this study,
we shall give another proof of the inverse-time formula.
%
We then try to derive a similar result
for the optimal updating magnitude
for updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
It turns out that the optimal updating magnitude
depends nontrivially on the simulation length
and other parameters in the latter case.
%
We also shall show that
the single-bin scheme used by the WL algorithm
is best in asymptotic convergence,
although practical simulations may not need
to reach this regime.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

\subsubsection{\label{twolimits}
Asymptotically optimal schedule}

The asymptotic error given by Eq. \eqref{eq:error_asym1}
can be rewritten as
%
\begin{equation}
  \Err_A
  =
  \int_0^t
    {\mathcal L} \bigl[ q(\tau)\bigr]
    \, d\tau,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(\tau) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl( q(\tau) \bigr)
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl( q(\tau) \bigr)
  \; \dot q^2( \tau ).
  \label{eq:error_asym_lagr}
\end{align}
%
To minimize the functional,
we notice that the Lagrangian given by Eq. \eqref{eq:error_asym_lagr}
does not explicitly depend on time, $t$.
%
Thus, the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved, and in this case,
the Hamiltonian and the Lagrangian coincide.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(\tau) \bigr]
  \;
  \dot q^2(\tau)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating integrate the square root of the left-hand side,
%
\begin{equation}
  \int_0^{ q(\tau) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, d\tau
  ,
  \label{eq:q_opt}
\end{equation}
%
we get an implicit equation for the optimal $\alpha(t')$
via $q(\tau) = \int_0^\tau \alpha(t') \, dt'$.

Although the explicit solution of Eq. \eqref{eq:q_opt} is difficult,
we can obtain an expression regarding $\alpha(\tau)$
by differentiating Eq. \eqref{eq:Lagrangian_const},
which yields
%
\begin{equation}
  \frac{ d      }
       { d \tau }
  \left(
    \frac{       1        }
         { \alpha( \tau ) }
  \right)
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl( q(\tau) \bigr)
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl( q(\tau) \bigr)
  }
  .
  \label{eq:dinvadt}
\end{equation}
%

Consider examine two limiting cases.
%
First, during the beginning of a long simulation,
we have $\tau \ll t$ and $q(t) - q(\tau) \to \infty$,
the sums on the right-hand side of Eq. \eqref{eq:dinvadt}
is dominated by the term with the largest
$u_k = e^{ \lambda_k \, [q(\tau) - q(t)] }$,
or, equivalently, with the smallest $\lambda_k$.
Then
\begin{equation}
  \frac{ d      }
       { d \tau }
  \left(
    \frac{       1        }
         { \alpha( \tau ) }
  \right)
  =
  \lambda_{ \min }
  ,
  \notag
  %\label{eq:dinvadt_limit1}
\end{equation}
%
and
%
\begin{equation}
  \alpha( \tau )
  =
  \frac{               1                  }
       { \lambda_{ \min } ( \tau + t_1 )  }
  .
  \label{eq:alpha_limit1}
\end{equation}
%

Second, near the end of simulation,
we have $\tau \to t$, $u_k \approx 1$,
%
\begin{equation}
  \frac{ d      }
       { d \tau }
  \left(
    \frac{       1        }
         { \alpha( \tau ) }
  \right)
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
  }
  =
  \bar \lambda,
  \notag
  %\label{eq:dinvadt_limit2}
\end{equation}
%
where,
$\bar \lambda$
is the average of $\lambda_k$ weighted by $\Gamma_k \, \lambda_k^2$,
and it tends to be dominated
by the largest $\lambda_k$.
%
Thus, in this regime, we have
%
\begin{equation}
  \alpha( \tau )
  =
  \frac{             1                }
       { \bar \lambda ( \tau + t_2 )  }
  .
  \label{eq:alpha_limit2}
\end{equation}
%
Thus, we may characterize the optimal schedule as
%
\begin{equation}
  \alpha(\tau)
  =
  \frac{                      1                          }
       { \lambda(\tau) \, \bigl[ \tau + t_0(\tau) \bigr] }
,
\notag
\end{equation}
%
where $\lambda(\tau)$ and $t_0(\tau)$ are two
slowly-varying functions,
with the former increases gradually from $\lambda_{\min}$
to $\bar \lambda$.

%Both Eqs. \eqref{eq:alpha_limit1} and \eqref{eq:alpha_limit2}
%resemble Eq. \eqref{eq:alpha_opt}.
